% !TEX root = ../thesis.tex

\chapter*{Acknowledgements}

This thesis would not have been possible without the help of many people.

For this reason, I would like to express my utmost gratitude to Prof. Dr. Ralph Engel and Prof. Dr. Guido Drexlin for inviting me to the institute for 
astroparticle physics, and for refereeing this work.

Thanks to Dr. Markus Roth who enabled me to travel to Argentina, where I could visit the observatory in person, and experience la vida argentina. More importantly
also thanks for his feedback and insights that he shared with me during the last year, and for being an excellent group leader.

Dr. David Schmidt and Dr. Darko Veberi\v{c} were always available for any question that I had, and played an instrumental role in shaping the analysis with their
thoughtful comments, and critical discussions. Dr. Schmidt especially continued to motivate me and helped solve any problems I encountered during my work, even 
while he was located on a completely different hemisphere with several timezones between us complicating communication.

One person that is criminally underrepresented in these acknowledgements is Sabine Bucher. I cannot thank her enough for all her tireless work and organizational
prowess. I am confident in saying that without her skills, the institute for astroparticle physics would not be as productive as it is. Thank you Sabine for all
your support in filling out numerous business trip applications, and organizing a large part of my trip to Argentina.

Finally, I want to express a heartfelt appreciation to all my office colleagues, Auger group members and IAP associates. Your questions, recommendations and 
general interest helped me realize the relevance of this work. All the moments we shared outside of working hours sensitized me to the importance of collaboration
